---
title: "Changes in microbial communities during international travel"
author: "JReceveur"
output: 
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    code_folding: hide
---
  
  ```{r setup, include=FALSE}
knitr::opts_chunk$set(fig.width=14, fig.height=10)
knitr::opts_chunk$set(fig.align="center")
```

To show the code used click either the code tab on the right side above each box or at the top of the document

``` {r import, message=FALSE, warning=FALSE}
library(vegan)
library(MASS)
library(ggplot2)
library(plyr)
library(dplyr)
library(magrittr)
library(scales)
library(grid)
library(reshape2)
library(phyloseq)
library(randomForest)
library(knitr)
library(ape)
library(ggpubr)
cbPalette <- c("#999999", "#E69F00", "#56B4E9", "#009E73", "#F0E442", "#0072B2", "#D55E00", "#000000","#CC79A7")
theme_set(theme_bw(base_size = 18)+theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank()))
biom=import_biom("C:\\Users\\Joe Receveur\\Documents\\Virtual Box\\BeninSwab\\Outputs\\BeninSwabR.biom",parseFunction= parse_taxonomy_greengenes)
meta=read.table("C:\\Users\\Joe Receveur\\Documents\\Virtual Box\\BeninSwab\\Metadata\\sample-metadata.txt",header=TRUE)
meta=sample_data(meta)
sample_names(meta)=meta$SampleID
physeq=merge_phyloseq(biom,meta) 
physeq <- subset_taxa(physeq, Family != "mitochondria" & Class != "Chloroplast")
GP= rarefy_even_depth(physeq, sample.size = 17500,
  rngseed = FALSE, replace = TRUE, trimOTUs = TRUE, verbose = TRUE)
GP
physeq=GP

sample_data(physeq)$Timepoint = factor(sample_data(physeq)$Timepoint, levels = c("DayOfDeparture","ArrivalBenin","DepartingBenin","ArrivalGeneva","DepartingGeneva","ArrivalHome","1WeekPost","2WeekPost")) #fixes x-axis labels

```

# Overview

Study Design: Samples were taken from four locations (Ears, Eyes, Nares, and Mouth) on four subjects during international travel to Benin and Switzerland over the course of two weeks and one week after returning.

Only Mouth and Nares samples have been sequenced from the following timepoints: DayOfDeparture, ArrivalBenin, DepartingBenin, ArrivalGeneva, DepartingGeneva, ArrivalHome, 1WeekPost, 2WeekPost

##Alpha diversity
```{r arichness, warning=FALSE}
plot_richness(physeq, x="Timepoint",color="Timepoint", measures=c("Simpson"))+geom_boxplot(aes(x=Timepoint, y=value, color=Timepoint), alpha=0.05)

```

##Taxonomic Composition  

```{r filter3levels, warning=FALSE}
GPr  = transform_sample_counts(physeq, function(x) x / sum(x) ) #transform samples based on relative abundance
GPrPhylum=tax_glom(GPr, "Phylum")
PhylumLevel = filter_taxa(GPrPhylum, function(x) mean(x) > 1e-3, TRUE) #filter out any taxa lower tha 0.1%
GPrFamily=tax_glom(GPr,"Family")
FamilyLevel = filter_taxa(GPrFamily,function(x) mean(x) > 1e-2, TRUE) #filter out any taxa lower tha 1%
GPrGenus=tax_glom(GPr,"Genus")
GenusLevel = filter_taxa(GPrGenus, function(x) mean(x) > 1e-2, TRUE) #filter out any taxa lower tha 1%
```

Phylum Level stacked bar graph
```{r phylumlevelplot,warning=FALSE}
df <- psmelt(PhylumLevel)
df$Abundance=df$Abundance*100
Trtdata <- ddply(df, c("Phylum", "Timepoint","Swab"), summarise,
                 N    = length(Abundance),
                 mean = mean(Abundance),
                 sd   = sd(Abundance),
                 se   = sd / sqrt(N)
)
cdataplot=ggplot(Trtdata, aes(x=Timepoint,y=mean))+geom_bar(aes(fill = Phylum),colour="black", stat="identity",position = "fill")+ facet_grid(~Swab, labeller=labeller(Sample_Type = labels))+xlab("Treatment")+ylab("Relative Abundance (> 0.1%)") + theme(axis.text.x = element_text(angle = 0, hjust = 0.5))+theme(axis.title.x=element_blank())+
  theme(axis.text.x = element_text(angle = 45, hjust = 1))+ scale_fill_manual(values=cbPalette)
cdataplot


```

#Mouth Taxa Plots

```{r TaxGlom}
physeq=subset_samples(physeq,Timepoint!="2WeekPost")
sample_data(physeq)$Timepoint =factor(sample_data(physeq)$Timepoint, levels = c("DayOfDeparture","ArrivalBenin","DepartingBenin","ArrivalGeneva","DepartingGeneva","ArrivalHome","1WeekPost","2WeekPost"))
LInt=subset_samples(physeq,Swab=="Mouth")
LInt=subset_samples(LInt,Timepoint!="2WeekPost")
GPr  = transform_sample_counts(LInt, function(x) x / sum(x) ) #transform samples based on relative abundance
GPrPhylum=tax_glom(GPr, "Phylum")
PhylumLevel = filter_taxa(GPrPhylum, function(x) mean(x) > 1e-3, TRUE) #filter out any taxa lower tha 0.1%
GPrFamily=tax_glom(GPr,"Family")
FamilyLevel = filter_taxa(GPrFamily, function(x) mean(x) > 1e-2, TRUE) #filter out any taxa lower tha 1%
GPrGenus=tax_glom(GPr,"Genus")
GenusLevel = filter_taxa(GPrGenus, function(x) mean(x) > 1e-2, TRUE) #filter out any taxa lower tha 1%

```


###Phylum level

Stars (*) on plot represent unadjusted p values for the test for each taxa (Wilcox or KW test depending on number of groups), pairwise adjusted p values are in the table below (FDR adjustment)
Table is filtered to only show taxa with a adjusted p value below 0.1

```{r, warning=FALSE}
df <- psmelt(PhylumLevel)
df$Abundance=df$Abundance*100
Trtdata <- ddply(df, c("Phylum", "Timepoint"), summarise,
                 N    = length(Abundance),
                 mean = mean(Abundance),
                 sd   = sd(Abundance),
                 se   = sd / sqrt(N)
)
```

```{r}
p <- ggbarplot(df, x = "Timepoint", y = "Abundance",add = c("mean_se"),#"mean_se"
               color = "black", palette = "cbPalette", facet.by="Phylum",
               line.color = "gray", line.size = 0.4, short.panel.labs = TRUE, p.adjust.method = "fdr", fill= "Timepoint") + stat_compare_means(aes(group = Timepoint), label = "..p.signif..",label.y = 7) 

p+ theme(axis.text.x = element_text(angle = 45, hjust = 1))+ylab("Relative abundance (> 0.1%)")+ theme(legend.position="none")

Means=compare_means(Abundance ~ Timepoint, data = df, 
              group.by = "Phylum", p.adjust.method = "fdr")
#head(Means)
keeps <- c("Phylum","group1","group2","p.format","p.adj","method","p.signif")
keeps=Means[keeps]
#keeps


test3 <- list('Phylum'= keeps$Phylum,'group1'=keeps$group1,'group2'= keeps$group2 ,'p'=keeps$p.format,'p.adj'=keeps$p.adj,p.signif=keeps$p.signif,'Method'=keeps$method)
test3= as.data.frame(test3)
#test3
FilteredResults<-test3[!(test3$p.adj>0.1),]            
FilteredResults
```


###Family Level

```{r,warning=FALSE}
df <- psmelt(FamilyLevel)
df$Abundance=df$Abundance*100
Trtdata <- ddply(df, c("Family", "Timepoint"), summarise,
                 N    = length(Abundance),
                 mean = mean(Abundance),
                 sd   = sd(Abundance),
                 se   = sd / sqrt(N)
)
```

```{r}
p <- ggbarplot(df, x = "Timepoint", y = "Abundance",add = c("mean_se"),#"mean_se"
               color = "black", palette = "cbPalette", facet.by="Family",
               line.color = "gray", line.size = 0.4, short.panel.labs = TRUE, p.adjust.method = "bonferroni", fill= "Timepoint") + stat_compare_means(aes(group = Timepoint), label = "..p.signif..",label.y = 7) 

p+ theme(axis.text.x = element_text(angle = 45, hjust = 1))+ylab("Relative abundance (> 1%)")+ theme(legend.position="none")

Means=compare_means(Abundance ~ Timepoint, data = df, 
              group.by = "Family", p.adjust.method = "fdr")
#head(Means)
keeps <- c("Family","group1","group2","p.format","p.adj","method","p.signif")
keeps=Means[keeps]
#keeps


test3 <- list('Family'= keeps$Family,'group1'=keeps$group1,'group2'= keeps$group2 ,'p'=keeps$p.format,'p.adj'=keeps$p.adj,p.signif=keeps$p.signif,'Method'=keeps$method)
test3= as.data.frame(test3)
#test3
FilteredResults<-test3[!(test3$p.adj>0.1),]            
FilteredResults
```

###Genus level

```{r, warning=FALSE}
df <- psmelt(GenusLevel)
df$Abundance=df$Abundance*100
Trtdata <- ddply(df, c("Genus", "Timepoint"), summarise,
                 N    = length(Abundance),
                 mean = mean(Abundance),
                 sd   = sd(Abundance),
                 se   = sd / sqrt(N)
)
```

```{r}
p <- ggbarplot(df, x = "Timepoint", y = "Abundance",add = c("mean_se"),#"mean_se"
               color = "black", palette = "cbPalette", facet.by="Genus",
               line.color = "gray", line.size = 0.4, short.panel.labs = TRUE, p.adjust.method = "fdr", fill= "Timepoint") + stat_compare_means(aes(group = Timepoint), label = "..p.signif..",label.y = 7) 

p+ theme(axis.text.x = element_text(angle = 45, hjust = 1))+ylab("Relative abundance (> 1%)")+ theme(legend.position="none")

Means=compare_means(Abundance ~ Timepoint, data = df, 
              group.by = "Genus", p.adjust.method = "fdr")
#head(Means)
keeps <- c("Genus","group1","group2","p.format","p.adj","method","p.signif")
keeps=Means[keeps]
#keeps


test3 <- list('Genus'= keeps$Genus,'group1'=keeps$group1,'group2'= keeps$group2 ,'p'=keeps$p.format,'p.adj'=keeps$p.adj,p.signif=keeps$p.signif,'Method'=keeps$method)
test3= as.data.frame(test3)
#test3
FilteredResults<-test3[!(test3$p.adj>0.1),]            
FilteredResults
```

#Nares taxa plots

```{r}
LInt=subset_samples(physeq,Swab=="Nares")
LInt=subset_samples(LInt,Swab!="2WeekPost")
GPr  = transform_sample_counts(LInt, function(x) x / sum(x) ) #transform samples based on relative abundance
GPrPhylum=tax_glom(GPr, "Phylum")
PhylumLevel = filter_taxa(GPrPhylum, function(x) mean(x) > 1e-2, TRUE) #filter out any taxa lower tha 1%
GPrFamily=tax_glom(GPr,"Family")
FamilyLevel = filter_taxa(GPrFamily, function(x) mean(x) > 1e-2, TRUE) #filter out any taxa lower tha 1%
GPrGenus=tax_glom(GPr,"Genus")
GenusLevel = filter_taxa(GPrGenus, function(x) mean(x) > 1e-2, TRUE) #filter out any taxa lower tha 1%

```


###Phylum level
Stars (*) on plot represent unadjusted p values for the test for each taxa (Wilcox or KW test depending on number of groups), pairwise adjusted p values are in the table below (FDR adjustment)
Table is filtered to only show taxa with a adjusted p value below 0.1

```{r, warning=FALSE}
df <- psmelt(PhylumLevel)
df$Abundance=df$Abundance*100
Trtdata <- ddply(df, c("Phylum", "Timepoint"), summarise,
                 N    = length(Abundance),
                 mean = mean(Abundance),
                 sd   = sd(Abundance),
                 se   = sd / sqrt(N)
)
```

```{r, warning=FALSE}
p <- ggbarplot(df, x = "Timepoint", y = "Abundance",add = c("mean_se"),#"mean_se"
               color = "black", palette = "cbPalette", facet.by="Phylum",
               line.color = "gray", line.size = 0.4, short.panel.labs = TRUE, p.adjust.method = "fdr", fill= "Timepoint") + stat_compare_means(aes(group = Timepoint), label = "..p.signif..",label.y = 7) 

p+ theme(axis.text.x = element_text(angle = 45, hjust = 1))+ylab("Relative abundance (> 1%)")+ theme(legend.position="none")

Means=compare_means(Abundance ~ Timepoint, data = df, 
              group.by = "Phylum", p.adjust.method = "fdr")
#head(Means)
keeps <- c("Phylum","group1","group2","p.format","p.adj","method","p.signif")
keeps=Means[keeps]
#keeps


test3 <- list('Phylum'= keeps$Phylum,'group1'=keeps$group1,'group2'= keeps$group2 ,'p'=keeps$p.format,'p.adj'=keeps$p.adj,p.signif=keeps$p.signif,'Method'=keeps$method)
test3= as.data.frame(test3)
#test3
FilteredResults<-test3[!(test3$p.adj>0.1),]            
FilteredResults
```


###Family Level


```{r,warning=FALSE}
df <- psmelt(FamilyLevel)
df$Abundance=df$Abundance*100
Trtdata <- ddply(df, c("Family", "Timepoint"), summarise,
                 N    = length(Abundance),
                 mean = mean(Abundance),
                 sd   = sd(Abundance),
                 se   = sd / sqrt(N)
)
```

```{r,warning=FALSE}
p <- ggbarplot(df, x = "Timepoint", y = "Abundance",add = c("mean_se"),#"mean_se"
               color = "black", palette = "cbPalette", facet.by="Family",
               line.color = "gray", line.size = 0.4, short.panel.labs = TRUE, p.adjust.method = "bonferroni", fill= "Timepoint") + stat_compare_means(aes(group = Timepoint), label = "..p.signif..",label.y = 7) 

p+ theme(axis.text.x = element_text(angle = 45, hjust = 1))+ylab("Relative abundance (> 1%)")+ theme(legend.position="none")

Means=compare_means(Abundance ~ Timepoint, data = df, 
              group.by = "Family", p.adjust.method = "fdr")
#head(Means)
keeps <- c("Family","group1","group2","p.format","p.adj","method","p.signif")
keeps=Means[keeps]
#keeps


test3 <- list('Family'= keeps$Family,'group1'=keeps$group1,'group2'= keeps$group2 ,'p'=keeps$p.format,'p.adj'=keeps$p.adj,p.signif=keeps$p.signif,'Method'=keeps$method)
test3= as.data.frame(test3)
#test3
FilteredResults<-test3[!(test3$p.adj>0.1),]            
FilteredResults
```

###Genus level

```{r, warning=FALSE}
df <- psmelt(GenusLevel)
df$Abundance=df$Abundance*100
Trtdata <- ddply(df, c("Genus", "Timepoint"), summarise,
                 N    = length(Abundance),
                 mean = mean(Abundance),
                 sd   = sd(Abundance),
                 se   = sd / sqrt(N)
)
```

```{r,warning=FALSE}
p <- ggbarplot(df, x = "Timepoint", y = "Abundance",add = c("mean_se"),#"mean_se"
               color = "black", palette = "cbPalette", facet.by="Genus",
               line.color = "gray", line.size = 0.4, short.panel.labs = TRUE, p.adjust.method = "fdr", fill= "Timepoint") + stat_compare_means(aes(group = Timepoint), label = "..p.signif..",label.y = 7) 

p+ theme(axis.text.x = element_text(angle = 45, hjust = 1))+ylab("Relative abundance (> 0.1%)")+ theme(legend.position="none")

Means=compare_means(Abundance ~ Timepoint, data = df, 
              group.by = "Genus", p.adjust.method = "fdr")
#head(Means)
keeps <- c("Genus","group1","group2","p.format","p.adj","method","p.signif")
keeps=Means[keeps]
#keeps


test3 <- list('Genus'= keeps$Genus,'group1'=keeps$group1,'group2'= keeps$group2 ,'p'=keeps$p.format,'p.adj'=keeps$p.adj,p.signif=keeps$p.signif,'Method'=keeps$method)
test3= as.data.frame(test3)
#test3
FilteredResults<-test3[!(test3$p.adj>0.1),]            
FilteredResults
```


#PERMANOVA/Ordination results

-All ordinations were conducted using Jensen-Shannon distance

-Ellipses represent 95% CI for the mean of each group

##PCoA 

```{r, warning=FALSE,message=FALSE}
ord=ordinate(physeq,"PCoA", "jsd")
ordplot=plot_ordination(physeq, ord,"samples", color="Timepoint")+geom_point(size=4)+scale_colour_manual(values=cbPalette)+scale_fill_manual(values=cbPalette)
ordplot+ stat_ellipse(type= "norm",geom = "polygon", alpha = 1/4, aes(fill = Timepoint))+ theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank())+
  facet_wrap(~Swab)#+ theme(legend.justification=c(1,0), legend.position=c(1,0))
```

```{r,warning=FALSE, message=FALSE}

ord=ordinate(physeq,"PCoA", "jsd")
ordplot=plot_ordination(physeq, ord,"samples", color="Swab")+geom_point(size=4)+scale_colour_manual(values=cbPalette)+scale_fill_manual(values=cbPalette)
ordplot+ stat_ellipse(type= "norm",geom = "polygon", alpha = 1/4, aes(fill = Swab))+ theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank())+
  facet_wrap(~Timepoint)#+ theme(legend.justification=c(1,0), legend.position=c(1,0))


```


##PERMANOVAs

##Swab type X Timepoint
See QIIME outputs for other distance metrics

``` {r,warning=FALSE}
GPdist=phyloseq::distance(physeq, "jsd")
```

```{r}
adonis(GPdist ~ Swab*Timepoint, as(sample_data(physeq), "data.frame"))
```
